### 2024-06-28 manuscript/manuscript.tex
uniformize species name in the final data set

\section*{Introduction}
The composition of ecological communities, defined as the abundance of each species within a given community, is fundamental for understanding patterns and processes in community ecology. Community composition can help to detect spatial patterns linked to environmental variations \citep{kemp1990}, assess temporal trends of different groups of species after disturbances \citep{philippi1998, magurran2007}, and understand food web structures \citep{cohen2003}. Additionally, community composition is essential for modeling the dynamics of ecological communities. Dynamic community modelling allows addressing important issues and questions in ecology, such as: determining the relative strength of top-down versus bottom-up forces in communities \citep{krebs2003,legagneux2014}, assessing the ecological resilience of communities under climate change \citep{griffith2019} and evaluating the cascading effects of invasive species in food web \citep{david2017, goto2020}. Dynamic community modelling can also be applied to address practical challenges, including fishery management (Plagányi, 2007) and the planning of protected areas \citep{okey2004, dahood2020}.

@@

The composition of ecological communities is influenced by various factors acting at different temporal and spatial scales, leading to noisy data and emphasizing the need for long-term data sets \citep{magurran2010, lindenmayer2012}. Species abundances are influenced by stochastic effects \citep{hubbell2001}, environmental changes (e.g., climate warming), and species interactions, contributing to data variability. For instance, a community composition could be driven simultaneously by the intra-annual seasonal variations in the environment, the multi-year dynamics of small mammals cycles and a relatively slow but directional climate warming. Therefore, to untangle the relative effects of diverse abiotic and biotic factors on community composition, long-term data sets are required \citep{magurran2010, lindenmayer2012}.

@@

\subsection*{Objective}
Our objective is to provide readily accessible, long-term time series of annual abundances of all vertebrate species within the Arctic terrestrial community of Bylot Island during the breeding season (May to August). This includes both breeding and non-breeding individuals that stay in the study area for a significant period of time, but excludes non-breeding individuals that stop for only a few days during their migration. Additionally, our focus extends to estimating abundances at the landscape scale, enabling the study of community and ecosystem dynamics, trophic interactions and the impacts of global changes on high-latitude environments.
\newpage

@@

The study area is divided into 9 zones based on the sampling method and the level of field effort applied in each zone (\textbf{Figure \ref{figure:zones}}). Long-term monitoring of the community began in the 1990s in the Qarlikturvik valley \citep{gauthier2013, gauthier2024a}, which represents the zone of the study area with the highest annual sampling effort. Within the Qarlikturvik valley, the sampling is concentrated on the southern side of the glacial river, where the main research infrastructure is situated. Another zone with extensive sampling efforts is Camp 2, located at the core of the snow goose colony, where the primary focus is to monitor snow goose nests. However, nests of many other avian species are also monitored within and around the colony in this zone. Camp 3, Pointe Dufour, Goose Point, and Malaview are zones where intensive sampling efforts are conducted annually, albeit for a relatively brief period (approximately one week) during the breeding season of most species \citep{gauthier2024a}. The upland zones in the study area (defined as areas approximately 300 meters above sea level or more) are the Black Plateau, Southern Plateau, and Camp 3 Plateau. These zones are primarily visited to assess raptor nesting activity \citep{beardsell2016}. The zone between the Qarlikturvik valley and Camp 3 received very little sampling effort and is therefore excluded from the study area.

@@

\subsubsection{Systematic nest monitoring}
\subsubsection*{Pacific loon, red-throated loon, cackling goose, tundra swan and glaucous gull}
Since 2004, systematic searches of wetland areas has been conducted on the southern side of the glacial river in the Qarlikturvik Valley, and since 2017, in other zones of the study area. This sampling aimed to find all nests of the cackling goose and the glaucous gull. Nest locations of other large wetland-nesting species, including the tundra swan, the red-throated loon and the Pacific loon, were also noted, as these species nest in the same habitat \citep{duchesne2021,gauthier2024a}. Each year, all known or potential nesting sites were revisited. Observers detected nests by walking and scanning around ponds and lakeshores to identify any active nesting sites. These large species can be seen from a relatively long distance sitting on the nest or when flushing from the nest. Most of them (geese, swans and gulls) can also reveal their presence with alarm calls or nest defense displays. We are confident that nest detection probability was high for these species given the open landscape.

@@

\subsubsection*{Rough-legged hawk, peregrine falcon and common raven}
Peregrine falcons, rough-legged hawks and common ravens nest on cliffs, near ravines, and on large rocky outcrops and tend to reuse the same nesting sites from one year to the next \citep{beardsell2016}. Systematic monitoring of every known or potential nesting site has been carried out in the Qarlikturvik valley, Black plateau and Southern plateau since 2007 and throughout the study area since 2013 \citep{beardsell2016, gauthier2020avianpred}. Observers walked along ridges and scan surrounding areas from vantage points to detect nesting birds. These large species can be seen from a relatively long distance sitting on the nest or when flushing from the nest. They can also reveal their presence with alarm calls or nest defense displays. We are confident that nest detection probability was high for these species. Each year the observers use slightly different paths to sample the areas, but locate the nests in the same positions, which supports a high probability of detection for these species. Most nesting sites were located in the upland zones of the study area, which are the Black Plateau, Southern Plateau and Camp 3 Plateau.

@@

\subsubsection*{Long-tailed jaeger and parasitic jaeger}
Since 2004, observers have walked parallel transects spaced 400 meters apart, covering the entire southern side of the glacial river in the Qarlikturvik Valley (33 km\textsuperscript{2}; \textbf{Figure \ref{figure:qarlikturvik_valley}}), during the nesting period. The aim of those transects was to record nests of long-tailed jaegers, parasitic jaegers, and sandhill cranes. Observers listened for alarm calls to detect territorial birds, and then located nests by observing the birds returning to their nests from elevated vantage points. We consider the sampling to be systematic for long-tailed and parasitic jaeger, since those species tend to leave their nest relatively far from the observer to perform mobbing behavior, and thus increasing their detection probability. We do not consider the sampling to be systematic for sandhill cranes as they only display defensive behaviors near their nests at relatively short distances.

@@

\subsubsection*{Lapland longspur and Baird's sandpiper}
Since 2005, nests of passerines and sandpipers have been extensively monitored across an 8 km\textsuperscript{2} (4x2 km) area in the Qarlikturvik valley. We considered the sampling to be most systematic within a core 2 km\textsuperscript{2} (2x1 km) plot in this area (\textbf{Figure \ref{figure:qarlikturvik_valley}}). An observer conducted systematic searches of this plot during the entire breeding season to locate and monitor as many passerine and shorebird nests as possible. Assuming this observer can detect all nests within 5 or 10 meters, this single observer sampled annually at minimum an area of 0.72 \textpm{0.12} or 1.09 \textpm{0.17} km\textsuperscript{2} (n= 3). Additionally, several other observers conducting related field work in the same zone reported all passerine and shorebird nests found opportunistically. We excluded relatively large water bodies (0.26 km\textsuperscript{2}) to calculate nest density in the plot due to the presence of a large lake, which leaves an area of 1.74 km\textsuperscript{2} available for nesting.

@@

\subsection{Observation of individuals}
\subsubsection{Vertebrate count transects}
From 2010 to 2023, observers walked 500-meter linear transects where all vertebrate individuals observed within 150 meters on either side were counted (146 to 320 transects per year). Transects were distributed across all lowland zones of the study area, typically in mesic habitats, and were carried out during the nesting period (between June 21 and July 14; \citet{duchesne2021}; \textbf{Figure \ref{figure:transects}}). Furthermore, specifically for American golden-plovers, we measured the distance of each observed individual to the transect path.

@@

\subsubsection{Arctic fox movement tracking}
Between 2008 and 2016, 109 Arctic foxes were fitted with Argos Platform Transmitter Terminals mounted on collars \citep{lai2015,christin2015, dulude2023}. Foxes were captured between May and August across the study area, within and outside the goose colony \citep{dulude2023}. Sampling of animal locations was set for an interval of 1 or 2 days and only locations between May 1 and October 30 were retained \citep{dulude2023}.

@@

\subsection{Species body mass}
All vertebrate individuals captured for marking purposes were systematically weighed (snow goose (unpublished data), snowy owl \citep{therrien2012, robillard2018}, American-golden plovers\citep{lamarre2021},  common-ringed plovers \citep{leandri2019}, other shorebirds (Bêty, unpublished data), glaucous gulls \citep{gauthier2015}, long-tailed jaeger \citep{seyer2019}, parasitic jaegers (Gauthier and Therrien, unpublished data), Lapland longspurs (Gauthier and Bêty, unpublished data), lemmings \citep{gauthier2020lemmings}, ermine (Bilodeau and Bolduc, unpublished data) and Arctic foxes \citep{lai2015}). Mean body mass of vertebrate species was determined in priority with measurements from Bylot Island. When not available, we extracted mean body mass from the literature \citep{wilman2014}.

@@

R software (version 4.3.2) was used to process the data on a linux-gnu machine. Most of data manipulation was carried out with the \textit{dplyr} \citep{dplyr} and \textit{sf} packages \citep{sf}. All codes are available in the Dryad repository XXXXXXXXXXXXXX.
\newpage

@@

We observed a relatively strong correlation between the nest density of glaucous gulls in the Qarlikturvik valley and the nest density across the entire study area (R\textsuperscript{2}= 0.84, p = 0.16, n= 4). Consequently, we estimated the density of glaucous gulls at the scale of the study area between 2004 and 2016 based on the nest density in the Qarlikturvik valley ($y = 0.12406x + 0.13775$). However, we did not observe such strong relationships for loons and swans and thus we did not extend the time series. Regarding cackling geese, we observed signs of an exponential increase over time based on the annual number of nests found in various zones of the study area  and the indices of relative abundance (i.e., daily incidental observations and transects). We thus fitted an exponential model using the number of nests found annually over two distinct periods: in 1996 when the first nest was discovered, and then from 2017 to 2023 when sampling effort was systematic across the whole study area (\textbf{Figure \ref{figure:cackling}}). We used the fitted model to estimate abundance between 1996 and 2016 when monitoring was less systematic, which could potentially underestimate observed abundance as seen on \textbf{Figure \ref{figure:cackling}}. We multiplied nest density by two to obtain the abundance.

>>>>

\section*{Introduction}
The composition of ecological communities, defined as the abundance of each species within a given community, is fundamental for understanding patterns and processes in community ecology. Variations in community composition can help to detect spatial patterns linked to environmental variations \citep{kemp1990}, assess temporal trends of different groups of species after disturbances \citep{philippi1998, magurran2007}, and understand food web structures \citep{cohen2003}. Additionally, community composition is essential for modeling the dynamics of ecological communities. Dynamic community modelling allows addressing important issues and questions in ecology, such as: determining the relative strength of top-down versus bottom-up forces in communities \citep{krebs2003,legagneux2014}, assessing the ecological resilience of communities under climate change \citep{griffith2019} and evaluating the cascading effects of invasive species in food web \citep{david2017, goto2020}. Dynamic community modelling can also be applied to address practical challenges, including fishery management (Plagányi, 2007) and the planning of protected areas \citep{okey2004, dahood2020}.

@@

The composition of ecological communities is influenced by various factors acting at different temporal and spatial scales, leading to noisy data and emphasizing the need for long-term data sets \citep{magurran2010, lindenmayer2012}. Species abundances are influenced by stochastic effects \citep{hubbell2001}, environmental changes (e.g., climate warming), and species interactions, contributing to data variability. For instance, the composition of a community could be driven simultaneously by the intra-annual seasonal variations in the environment, the multi-year cyclic dynamics of an important prey or predator and a relatively slow but directional climate warming. Therefore, long-term data series are required to untangle the relative effects of diverse abiotic and biotic factors on community composition \citep{magurran2010, lindenmayer2012}.

@@

\subsection*{Objective}
Our main objective is to provide readily accessible, long-term time series of annual abundances of all vertebrate species within the Arctic terrestrial community of Bylot Island during the breeding season (May to August). This includes both breeding and non-breeding individuals that stay in the study area for a significant period of time, and excludes non-breeding individuals that stop for only a few days during their migration. Our focus extends to estimating abundances at the landscape scale, enabling the study of community and ecosystem dynamics, trophic interactions and the impacts of global changes on high-latitude environments. Additionally, we aim to provide the average body mass for each species in the community, enabling the conversion of abundances into biomasses.
\newpage

@@

The study area is divided into 9 zones based on the sampling method and the level of field effort applied in each zone (\textbf{Figure \ref{figure:zones}}). Long-term monitoring of the community began in the 1990s in the Qarlikturvik valley \citep{gauthier2013, gauthier2024a}, which represents the zone of the study area with the highest annual sampling effort. Within the Qarlikturvik valley, the sampling is concentrated on the southern side of the glacial river, where the main research infrastructure is located. Another zone with extensive sampling efforts is Camp 2, located at the core of the snow goose colony, where the primary focus is to monitor snow goose nests. However, nests of many other avian species are also monitored within and around the colony in this zone. Camp 3, Pointe Dufour, Goose Point, and Malaview are zones where intensive sampling efforts are conducted annually, albeit for a relatively brief period (approximately one week) during the breeding season of most species \citep{gauthier2024a}. The upland zones in the study area (defined as areas approximately 300 meters above sea level or more) are the Black Plateau, Southern Plateau, and Camp 3 Plateau. These zones are primarily visited to assess raptor nesting activity \citep{beardsell2016}. The zone between the Qarlikturvik valley and Camp 3 received very little sampling effort and is therefore excluded from the study area.

@@

\subsubsection{Systematic nest monitoring}
\subsubsection*{Pacific loon, red-throated loon, cackling goose, tundra swan and glaucous gull}
Since 2004, systematic searches of wetland areas have been conducted on the southern side of the glacial river in the Qarlikturvik Valley, and since 2017, in other zones of the study area. This sampling aimed to find all nests of the cackling goose and the glaucous gull. Nest locations of other large wetland-nesting species, including the tundra swan, the red-throated loon and the Pacific loon, were also noted, as these species nest in the same habitat \citep{duchesne2021,gauthier2024a}. Each year, all known or potential nesting sites were revisited. Observers detected nests by walking and scanning around ponds and lakeshores to identify any active nesting sites. These large species can be seen from a relatively long distance sitting on the nest or when flushing from the nest. Most of them (geese, swans and gulls) can also reveal their presence with alarm calls or nest defense displays. We are confident that nest detection probability was high for these species given the open landscape.

@@

\subsubsection*{Rough-legged hawk, peregrine falcon and common raven}
Peregrine falcons, rough-legged hawks and common ravens nest on cliffs, near ravines, and on large rocky outcrops and tend to reuse the same nesting sites from one year to the next \citep{beardsell2016}. Systematic monitoring of every known or potential nesting site has been carried out in the Qarlikturvik valley, Black plateau and Southern plateau since 2007 and throughout the study area since 2013 \citep{beardsell2016, gauthier2020avianpred}. Observers walked along ridges and scan surrounding areas from vantage points to detect nesting birds. These large species can be seen from a relatively long distance sitting on the nest or when flushing from the nest. They can also reveal their presence with alarm calls or nest defense displays. We are confident that nest detection probability was high for these species. Each year the observers use slightly different paths to sample the areas, but locate the nests in the same positions, which supports a high probability of detection for these species. Most nesting sites were located in the upland zones of the study area, which include the Black Plateau, Southern Plateau and Camp 3 Plateau.

@@

\subsubsection*{Long-tailed jaeger and parasitic jaeger}
Since 2004, observers have walked parallel transects spaced 400 meters apart, covering the entire southern side of the glacial river in the Qarlikturvik Valley (33 km\textsuperscript{2}; \textbf{Figure \ref{figure:qarlikturvik_valley}}), during the nesting period. The aim of those transects was to record nests of long-tailed jaegers, parasitic jaegers, and sandhill cranes. Observers listened for alarm calls to detect territorial birds, and then located nests by observing the birds returning to their nests from elevated vantage points. We consider the sampling to be systematic for long-tailed and parasitic jaeger, since those species tend to leave their nest relatively far from the observer to perform mobbing behavior, and thus increasing their detection probability. We do not consider the sampling to be systematic for sandhill cranes as they only display defensive behaviors near their nests at relatively short distances (see opportunistic nest monitoring below).

@@

\subsubsection*{Lapland longspur and Baird's sandpiper}
Since 2005, nests of passerines and sandpipers have been extensively monitored across an 8 km\textsuperscript{2} (4x2 km) area in the Qarlikturvik valley. We considered the sampling to be most systematic within a core 2 km\textsuperscript{2} (2x1 km) plot in this area (\textbf{Figure \ref{figure:qarlikturvik_valley}}). We excluded relatively large water bodies (0.26 km\textsuperscript{2}) to calculate nest density in the plot due to the presence of a large lake, which leaves an area of 1.74 km\textsuperscript{2} available for nesting. An observer conducted systematic searches of this plot during the entire breeding season to locate and monitor as many passerine and shorebird nests as possible. Assuming the observer can detect all nests within a 5 or 10 meter radius, analysis of daily GPS tracks shows that the observer covered a minimum area of 0.72 \textpm{0.12} (5 m) or 1.09 \textpm{0.17} km\textsuperscript{2} (10 m) of the core area annually (n= 3 years). Additionally, several other observers conducting related field work in the same zone reported all passerine and shorebird nests found opportunistically.

@@

\subsection{Observation of individuals}
\subsubsection{Vertebrate count transects}
From 2010 to 2023, observers walked 500-meter linear transects where all vertebrate individuals observed within 150 meters on either side were counted (146 to 320 transects per year). Transects were distributed across all lowland zones of the study area, typically in mesic habitat, and were carried out during the nesting period (between June 21 and July 14; \citet{duchesne2021}; \textbf{Figure \ref{figure:transects}}). Furthermore, specifically for American golden-plovers, we measured the distance of each observed individual to the transect path.

@@

\subsubsection{Arctic fox movement tracking}
In order to assess fox abundance based on the size of their home range, 109 Arctic foxes were fitted with Argos Platform Transmitter Terminals mounted on collars between 2008 and 2016 \citep{lai2015,christin2015, dulude2023}. Foxes were captured between May and August across the study area, within and outside the goose colony \citep{dulude2023}. Sampling of animal locations was set for an interval of 1 or 2 days and only locations between May 1 and October 30 were retained \citep{dulude2023}.

@@

\subsection{Species body mass}
All vertebrate individuals captured for marking purposes were systematically weighed (snow goose (unpublished data), snowy owl \citep{therrien2012, robillard2018}, American-golden plovers \citep{lamarre2021},  common-ringed plovers \citep{leandri2019}, other shorebirds (Bêty, unpublished data), glaucous gulls \citep{gauthier2015}, long-tailed jaeger \citep{seyer2019}, parasitic jaegers (Gauthier and Therrien, unpublished data), Lapland longspurs (Gauthier and Bêty, unpublished data), lemmings \citep{gauthier2020lemmings}, ermine (Bilodeau and Bolduc, unpublished data) and Arctic foxes \citep{lai2015}). Mean body mass of vertebrate species was determined in priority with measurements from Bylot Island. When not available, we extracted mean body mass from the literature \citep{wilman2014}.

@@

R software (version 4.3.2) was used to process the data on a linux-gnu machine. Most of data manipulation was carried out with the \textit{dplyr} \citep{dplyr} and \textit{sf} packages \citep{sf}. All codes are available in the following repository \url{https://github.com/chaireBioNorth/BYLOT_species_abundance_dataset.git}.
\newpage

@@

We observed a relatively strong correlation between the nest density of glaucous gulls in the Qarlikturvik valley and the nest density across the entire study area (R\textsuperscript{2}= 0.84, p = 0.16, n= 4). Consequently, we estimated the density of glaucous gulls at the scale of the study area between 2004 and 2016 based on the nest density in the Qarlikturvik valley ($y = 0.12406x + 0.13775$). However, we did not observe such strong relationships for loons and swans and thus we did not extend the time series. Regarding cackling geese, we observed signs of an exponential increase over time based on the annual number of nests found in various zones of the study area. We thus fitted an exponential model using the number of nests found annually over two distinct periods: in 1996 when the first nest was discovered, and then from 2017 to 2023 when sampling effort was systematic across the whole study area (\textbf{Figure \ref{figure:cackling}}). We used the fitted model to estimate abundance between 1996 and 2016 when monitoring was less systematic, which could potentially underestimate observed abundance as seen on \textbf{Figure \ref{figure:cackling}}. We multiplied nest density by two to obtain the abundance.

### 2024-06-28 manuscript/tables/table_summary_methods.tex
uniformize species name in the final data set

% Tue Jun 25 10:25:27 2024

@@

\caption{Summary of the methods and spatial scales used to estimate the density of each vertebrate species on Bylot Island, and how abundance was extrapoated at the landscape scale.} 

>>>>

% Fri Jun 28 12:52:36 2024

@@

\caption{Summary of the methods and spatial scales used to estimate the density of each vertebrate species on Bylot Island, and how abundance was extrapolated at the landscape scale. When sampling consist in relative abundance, we assumed that the ratios between relative and actual abundance are the same between the targeted species and a given reference species (i.e., similar detection probability). We therefore derived the absolute abundance of the targeted species from its relative abundance using the ratio between relative and absolute abundances of the reference species. Additionally, here the area of each sampling scale: two 11 ha grids (0.22 km2), 2x1km plot (2 km2) 4x2km plot (8 km2), south qarlikturvik (33 km2), camp 2 (80 km2), lowland (313 km2) and whole study area (389 km2).} 
